Raising the Role of Vocabulary Hubs for Semantic Data Interoperability in Dataspaces

### 2024-06-15 EBDVF-2024-elsevier/paper.qmd
updated EBDVF paper using elsevier template

---
title: Raising the Role of Vocabulary Hubs for Semantic Data Interoperability in Dataspaces

abstract: |
  Dataspaces are an important enabler for industrial sharing data (either commercially licensed or private).
  Europe is investing heavily into sectoral dataspaces, federation and orchestration platforms like SIMPL, Eclipse DSC, GXFS, etc.
  Still, dataspaces enable shared data access, but do not solve the data interoperability problem.
  For that, the consumer would like to see the data from different providers in a harmonized and semantically integrated form.
  The Vocabulary Hub service (part of the IDSA RAM) provides a repository for ontologies and vocabularies.
  We describe an approach of raising the role of the vocabulary hub to also allow richer metadata description (e.g. the meaning of every column in a tabular dataset),
  and binding semantic descriptions to ingested datasets, thus providing on-the-fly data semantization and easing data querying.
  This is achieved through the integration of two commercial semantic products (PoolParty and GraphDB),
  leveraging the partnership between the Semantic Web Company and Ontotext,
  and is being developed within the frame of the Digital Europe project UNDERPIN,
  with applications to refinery and wind farm data.

author:
  - name: Robert David
    email: robert.david@semantic-web.com
    orcid: 0000-0002-3244-5341
    affiliation: 
     name: The Semantic Web Company
     country: Austria
  - name: Vladimir Alexiev
    email: vladimir.alexiev@ontotext.com
    orcid: 0000-0001-7508-7428
  - name: Petar Ivanov
    email: petar.ivanov@ontotext.com
    orcid: 0000-0001-8448-1005
    affiliation: 
      name: Ontotext
      country: Bulgaria
acm-metadata:
  keywords:
    - dataspaces
    - semantic interoperability
    - semantic technologies
    - ontologies
    - vocabulary hub
    - oil and gas
    - renewable energy
    - refineries
    - windfarms
  final: true
  # acmart-options: sigconf, review
  # copyright-year: 2024
  # acm-year: 2024
  copyright: acmcopyright
  # doi: 
  # conference-acronym: 
  # conference-name: 
  # conference-date: 
  # conference-location: 
  # isbn:

format:
  acm-pdf:
    keep-tex: true 
bibliography: bibliography.bib

---

# Introduction
The European data economy heavily depends on the availability of data and the technological foundation to make use of it. 
Different industries and communities can benefit mutually by sharing data with each other, thereby supporting their digitization business goals. 
Machine-learning systems heavily rely on high volumes of high-quality training data. 
Examples are the well-known mobility dataspace and use cases like energy communities for running simulations. 
To cope with the challenge of data sharing, Dataspace approaches like the IDS RAM were introduced to not only solve the technical aspects of data sharing, 
but also to establish methods for data sovereignty so that dataspace participants can determine themselves how and when others can make use of their data.

For making use of shared data, interoperability is a crucial factor. 
While syntactic data exchange is well covered by clearly defined data formats, there is still the open challenge of semantic interoperability. 
IDS RAM is firmly based on semantic metadata for all aspects and actors of a dataspace. 
Furthermore, it can leverage semantic interoperability approaches through the Vocabulary Hub 
that stores ontologies and semantic thesauri for data descriptions and can dereference URIs to provide semantic details of data assets.

In this paper, we present our proposal for a technological solution of an IDS Vocabulary Hub 
built on the integration of the GraphDB semantic graph database and PoolParty Semantic Suite. 
We not only cover the core Vocabulary Hub functionalities and responsibilities in a dataspace, 
but also extend the idea towards a Semantic Layer for dataspaces, 
which provides even more services to easily and semi-automatically manage semantic interoperability. 
Thus we raise the role of the Vocabulary Hub to not only store semantic assets, 
but also to bind semantic descriptions dynamically to incoming data, 
therefore enabling richer discovery and semantic data integration between datasets.

@@

## GraphDB Graph Database
TODO

@@

Vocabulary Hub Services:
- RDF based: both GraphDB and PoolParty store and manage data based on the RDF data model. RDF is the basic building stone and underlying vocabulary definitions.
- Scalable vocabulary storage: cluster implementation for higher throughput and a fault tolerant architecture.
- Knowledge graph management: easy management of vocabularies using a rich web user interface and integration via Web APIs. 
  Vocabularies can be created and edited based on standards like RDFS, OWL and SKOS, and existing vocabularies can be reused and combined to fit specific dataspace needs.
- SPARQL service for semantic lookups: dereference URIs describing data assets to retrieve a semantic description of the entity from the graph database.

Semantic Layer Services support both unstructured and structured data assets:
- Unstructured data: semantic annotation services for documents based on concept tagging. 
  PoolParty’s KG-based NLP primarily uses SKOS taxonomies to identify concepts in documents, 
  but can be extended with OWL ontology elements (classes, properties) for increased expressiveness.
- Structured data: we support structured data harmonization methods, including semantic mapping and linking SKOS concepts
  - Ontology-Based Data Access (OBDA) is used for virtualization of relational data, i.e. accessing relational data as semantic data. 
    It is well suited for semantic querying of large amounts of data, while keeping the original relational data in place (NoETL approach). 
    Ontologies are leveraged to construct the OBDA mapping; 
    they are governed via the Semantic Layer and can harmonize data assets of various formats using a precise mapping. 
    Furthermore, it can be used to establish a fine-grained access control, thereby deciding for individual data elements if they should be exposed ot not.
  - RDB to RDF Mapping Language (R2RML) [@souripriyadasR2RMLRDBRDF2012]
    and RDF Mapping Language (RML) [@RMLIntroduction2024]
    are intended for data transformation processes within a dataspace, 
    where the original source is used as a basis to create an RDF representation (ETL approach).
  - GraphDB supports both OBDA and R2RML through its integration of ONTOP [@bagosiOntopFrameworkOntology2014]. 
    The same mappings can be used for NoETL virtual data access, or for ETL (to materialize the semantic data).
  - CSV on the Web (CSVW) [@swirrlCSVWStandards2024] is a set of W3C standards 
    that allow detailed semantic description of tabular data using a JSON "manifest" file.
  - SKOS concepts are used for structured data in several ways. 
    Concepts can be directly added to RDF data via semantic links (properties from an ontology), 
    they can be used to annotate specific literals of an RDF graph, 
    or they can be used to associate documents with (specific parts of) an RDF graph.
- Inference services: we provide metadata expansion services based on interlinked vocabularies in combination with ontologies. 
  This inference approach can be used with semantic tagging for unstructured data, as well as the structured data approaches.
  - Inference tagging is provided by a PoolParty tagging service in combination with an expansion query for implementing the inference. 
    The query can be configured to work for various scenarios.
- Vocabulary crosswalks are a data modeling approach using interlinks to relate:
  - Different vocabularies from similar or the same domain that cover different areas or aspects of the domain. 
    By relating concepts with each other using 1:1, 1:n and m:n relations, we can automatically expand queries and annotations by traversing the interlinked graph.
  - Different versions of the same vocabulary to automatically translate between those versions.
- Metadata inference based on interlinked vocabularies is similar to crosswalks, 
  but intended to represent different levels of abstraction, 
  so that we are able to infer from a more general vocabulary to a more specific vocabulary and vice versa. 
  For example, we can start with DCAT themes vocabularies and link to further vocabularies, which are more specific for certain areas or which go into more detail.

@@

The next steps include:
- Bringing it into actual dataspaces in practice, representing and integrating datasets from different dataspace participants.
- Implementing use cases for richer discoverability, harmonized querying and support for different content types of structured and unstructured data.
- Explore how ML can benefit regarding quality in practice when providing consolidated and cleaned data via dataspaces.
- Discuss how we can extend the IDS-RAM with services to improve the support for semantic interoperability provided by our solution.

# Acknowledgements
This work is partially supported by the Digital Europe programme project [UNDERPIN](https://underpinproject.eu/) (grant agreement 101123179)

# References

>>>>

---
title: Raising the Role of Vocabulary Hubs for Semantic Data Interoperability in Dataspaces
author:
  - name: Robert David
    email: robert.david@semantic-web.com
    orcid: 0000-0002-3244-5341
    affiliations: 
        - id: swc
          name: The Semantic Web Company
          country: Austria
  - name: Petar Ivanov
    email: petar.ivanov@ontotext.com
    orcid: 0000-0001-8448-1005
    affiliations: 
        - id: onto
          name: Ontotext
          country: Bulgaria
  - name: Vladimir Alexiev
    email: vladimir.alexiev@ontotext.com
    orcid: 0000-0001-7508-7428
    affiliations:
        - ref: onto
abstract: |
  Dataspaces are an important enabler for industrial sharing data (either commercially licensed or private).
  Europe is investing heavily into sectoral dataspaces, federation and orchestration platforms like SIMPL, Eclipse DSC, GXFS, etc.
  Still, dataspaces enable shared data access, but do not solve the data interoperability problem.
  For that, the consumer would like to see the data from different providers in a harmonized and semantically integrated form.
  The Vocabulary Hub service (part of the IDSA RAM) provides a repository for ontologies and vocabularies.
  We describe an approach of raising the role of the vocabulary hub to also allow richer metadata description (e.g. the meaning of every column in a tabular dataset),
  and binding semantic descriptions to ingested datasets, thus providing on-the-fly data semantization and easing data querying.
  This is achieved through the integration of two commercial semantic products (PoolParty and GraphDB),
  leveraging the partnership between the Semantic Web Company and Ontotext,
  and is being developed within the frame of the Digital Europe project UNDERPIN,
  with applications to refinery and wind farm data.
keywords: 
  - dataspaces
  - semantic interoperability
  - semantic technologies
  - ontologies
  - vocabulary hub
  - oil and gas
  - renewable energy
  - refineries
  - windfarms
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Workshop on Semantic Interoperability in Data Spaces 
      formatting: preprint
      model: 3p
      cite-style: super
---

# Introduction
The European data economy heavily depends on the availability of data and the technological foundation to make use of it. 
Different industries and communities can benefit mutually by sharing data with each other, thereby supporting their digitization business goals. 
Machine-learning systems heavily rely on high volumes of high-quality training data. 
Examples are the well-known mobility dataspace and use cases like energy communities for running simulations. 
To cope with the challenge of data sharing, Dataspace approaches like the IDS Reference Architecture Model (IDS RAM) were introduced to not only solve the technical aspects of data sharing, 
but also to establish methods for data sovereignty so that dataspace participants can determine themselves how and when others can make use of their data.

For making use of shared data, interoperability is a crucial factor. 
While syntactic data exchange is well covered by clearly defined data formats, there is still the open challenge of semantic interoperability [@woutervandenbergSemanticInteroperabilityData2024]. 
IDS RAM is firmly based on semantic metadata for all aspects and actors of a dataspace. 
Furthermore, it can leverage semantic interoperability approaches through the Vocabulary Hub [@IDSARAMVocabulary2024]
that stores ontologies and semantic thesauri for data descriptions and can dereference URIs to provide semantic details of data assets.

In this paper, we present our proposal for a technological solution of an IDS Vocabulary Hub 
built on the integration of the GraphDB semantic graph database and PoolParty Semantic Suite. 
We not only cover the core Vocabulary Hub functionalities and responsibilities in a dataspace, 
but also extend the idea towards a Semantic Layer for dataspaces, 
which provides even more services to easily and semi-automatically manage semantic interoperability. 
Thus we raise the role of the Vocabulary Hub [@stornebrinkmichielVocabularyHubSemantic2023] to not only store semantic assets, 
but also to bind semantic descriptions dynamically to incoming data, 
therefore enabling richer discovery and semantic data integration between datasets.

@@

## GraphDB Graph Database
Ontotext GraphDB is a highly efficient, scalable and robust graph database with RDF and SPARQL support.
It also offers data virtualization capabilities which enable direct access using SPARQL to relational databases, eliminating the need to replicate data.
It can perform semantic inferencing at scale, which enables deriving new semantic facts from existing facts.

GraphDB handles massive loads, queries, and inferencing in real time.
Reasoning and query evaluation are performed over a persistent storage layer. Loading, reasoning, and query evaluation are executed extremely fast even against huge ontologies and knowledge bases.
According to the [LDBC Semantic Publishing Benchmark](https://ldbcouncil.org/benchmarks/spb/), it is one of the most scalable OWL repositories currently available.

@@

Vocabulary Hub Services:

- RDF based: both GraphDB and PoolParty store and manage data based on the RDF data model. RDF is the basic building stone and underlying vocabulary definitions.
- Scalable vocabulary storage: cluster implementation for higher throughput and a fault tolerant architecture.
- Knowledge graph management: easy management of vocabularies using a rich web user interface and integration via Web APIs. 
  Vocabularies can be created and edited based on standards like RDFS, OWL and SKOS, and existing vocabularies can be reused and combined to fit specific dataspace needs.
- SPARQL service for semantic lookups: dereference URIs describing data assets to retrieve a semantic description of the entity from the graph database.

Semantic Layer Services support both unstructured and structured data assets:

- Unstructured data: semantic annotation services for documents based on concept tagging. 
  PoolParty’s KG-based NLP primarily uses SKOS taxonomies to identify concepts in documents, 
  but can be extended with OWL ontology elements (classes, properties) for increased expressiveness.
- Structured data: we support structured data harmonization methods, including semantic mapping and linking SKOS concepts
  - Ontology-Based Data Access (OBDA) is used for virtualization of relational data, i.e. accessing relational data as semantic data. 
    It is well suited for semantic querying of large amounts of data, while keeping the original relational data in place (NoETL approach). 
    Ontologies are leveraged to construct the OBDA mapping; 
    they are governed via the Semantic Layer and can harmonize data assets of various formats using a precise mapping [@andreselAdaptingOntologybasedData2024]. 
    Furthermore, it can be used to establish a fine-grained access control, thereby deciding for individual data elements if they should be exposed ot not.
  - RDB to RDF Mapping Language (R2RML) [@souripriyadasR2RMLRDBRDF2012]
    and RDF Mapping Language (RML) [@RMLIntroduction2024]
    are intended for data transformation processes within a dataspace, 
    where the original source is used as a basis to create an RDF representation (ETL approach).
  - GraphDB supports both OBDA and R2RML through its integration of ONTOP [@bagosiOntopFrameworkOntology2014]. 
    The same mappings can be used for NoETL virtual data access, or for ETL (to materialize the semantic data).
  - CSV on the Web (CSVW) [@swirrlCSVWStandards2024] is a set of W3C standards 
    that allow detailed semantic description of tabular data using a JSON "manifest" file.
  - SKOS concepts are used for structured data in several ways. 
    Concepts can be directly added to RDF data via semantic links (properties from an ontology), 
    they can be used to annotate specific literals of an RDF graph, 
    or they can be used to associate documents with (specific parts of) an RDF graph.
- Inference services: we provide metadata expansion services based on interlinked vocabularies in combination with ontologies. 
  This inference approach can be used with semantic tagging for unstructured data, as well as the structured data approaches.
  - Inference tagging is provided by a PoolParty tagging service in combination with an expansion query for implementing the inference. 
    The query can be configured to work for various scenarios.
- Vocabulary crosswalks are a data modeling approach using interlinks to relate:
  - Different vocabularies from similar or the same domain that cover different areas or aspects of the domain. 
    By relating concepts with each other using 1:1, 1:n and m:n relations, we can automatically expand queries and annotations by traversing the interlinked graph.
  - Different versions of the same vocabulary to automatically translate between those versions.
- Metadata inference based on interlinked vocabularies is similar to crosswalks, 
  but intended to represent different levels of abstraction, 
  so that we are able to infer from a more general vocabulary to a more specific vocabulary and vice versa. 
  For example, we can start with DCAT themes vocabularies and link to further vocabularies, which are more specific for certain areas or which go into more detail.

@@

The next steps include:

- Bringing it into actual dataspaces in practice, representing and integrating datasets from different dataspace participants.
- Implementing use cases for richer discoverability, harmonized querying and support for different content types of structured and unstructured data.
- Explore how ML can benefit regarding quality in practice when providing consolidated and cleaned data via dataspaces.
- Discuss how we can extend the IDS-RAM with services to improve the support for semantic interoperability provided by our solution.

# Acknowledgements
This work is partially supported by the Digital Europe programme project [UNDERPIN](https://underpinproject.eu/) (grant agreement 101123179).

# References {-}
